\vspace{-1cm} Model linearity breeds contempt: using Bayesian non-linear models to uncover broad macroecological patterns

### 2021-05-31 manuscript/manuscript.tex
New version of the draft

%Title paper
\title{\vspace{-1cm}
Model linearity breeds contempt: using Bayesian non-linear models to uncover broad macroecological patterns}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
% Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

\section*{Abstract}
Species' realized niches are classically pictured as bell-shaped probability distributions. These distributions, however, can actually take many different forms. For example, fat-tailed or skewed responses are very common across fields, as these can naturally emerge as a result of several ecological processes. While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies have shaped ecological communities. Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover such properties. These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies. With this approach, we are able to distil the shape of empirical plant distributions, which helps us tackle long-standing hypotheses regarding the way ecological communities are assembled across space. In particular, we studied the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients. Moreover, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical contexts. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

@@

%############# NEW INTRO ##############
%Bell shape is a good option, but it is unclear what the real shape is
One of the central goals of ecology is to understand the ways species are distributed across space and time. While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, abundant-centre-hypothesis}. In practice, there is a strong argument to be made in favour of assuming these to be bell shaped. Namely, if all that we are willing to assume about species' distributions is that these occupy finite geographic ranges, the most conservative statistical approach is to model their distribution as Gaussian (i.e.~the corresponding maximum entropy distribution; \citealt{frankCommonPatternsNature2009}). That said, there is currently no general agreement on the basic shape of species' realized niches. Indeed, many factors can play a role in defining their shape, and several natural processes can lead to non-normal distributions.

@@

%they can also reveal differences among species regarding the way they assemble along a gradient and broad macroecological patterns defining natural constraints shaping communities. 
%The general properties of species' distributions are valuable because they point to important ecological processes shaping species' realized niches. 
%For example, a relationship or lack thereof between range size and elevation would speak to classical ecological hypotheses (i.e.~Rapopor's rule;  \citealt{stevensElevationalGradientAltitudinal1992}) and could hint at the existence of general biogeographical constraints that shape species' distributions along gradients.
%Moreover, properties such as the skewness of species' distributions might capture other underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, bernat}. For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992, bernat}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients. This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments. That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}. Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem to solve. Indeed, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species along environmental gradients.
%Understanding the shape of species' realized niches and the extent to which these vary across species is therefore a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
%Understanding the shape of species' realized niches and the extent to which these vary across species is however challenging statistical problem. That is, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.
%With all of that being said, measuring these properties is a challenging statistical problem, and we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.

Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000}. These models are fundamental to the scientific community for predicting changes in species' geographic distributions and the effects of environmental disturbances. Such frameworks, however, commonly assume an underlying linear relationship between covariates (but see `semiparametric models'; \citealt{norbergComprehensiveEvaluationPredictive2019}). This is useful because it simplifies the optimization process but might not be ideal when studying and comparing the shape of species' distributions along environmental gradients. First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified. While this might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}. Second, the aforementioned structural constraints also limit our ability to include any prior information to our parameter estimates. Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}). Finally, \citet{huismanHierarchicalSetModels1993} proposed several non-linear structures to characterize several features of individual species' response curves. Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following these model structures, these were not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g. sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).

In this work, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy. In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns. Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or neophytes, might respond to the environment. More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

@@

Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 10 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). On the other hand, the information regarding species' ecological strategies represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology and life-history of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species introduction status (e.g.~identifying those species that are neophytes or archaeophytes) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades). We describe this information in more detail in the Supplementary Table 1.

%we independently consider those traits that describe growth and life strategies of the plants---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify $\sim120$ binary traits that characterize the physiology of species.

%comprising a total of 10 environmental indicator values with 10 types of associations for each of them (i.e. the $r$ associations defined above). Finally, we additionally independently consider species' range of variations for the such indicator values, also defining 10 `traits' with 3 types of associations (I need to see if I can added these to the analyses above).

%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19. Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland. There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.
%Based on 30 years (1961–1990) of meteorological data from national weather stations, fourteen climatic variables (Table S5) were calculated according to the method described in Zimmermann & Kienast44 and Zimmerman et al.55. A digital elevation model (DEM) was used to spatially interpolate the climatic data and to create topographic variables (Table S5). Additionally, we used a soil pH map based on Buri et al.43. All environmental data had a spatial resolution of 25 m, a resolution similar to the areas used to calculate the ecological indicator values (EIV; see below). This allowed a straightforward comparison of different sets of predictors composed from environmental variables and EIVs.

%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19. Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland. There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.

@@

\subsubsection*{Response curve}
To choose an appropriate response curve, we first need to agree on what we truly know about the system. Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance. While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions. Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution (Fig~\ref{fig:response}a). That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we can define species' responses to the environment as
\begin{equation}
\begin{split}
y_{ij} & \sim F\left(p_{ij}\right) \\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \gamma_{i} \left(x_{j}-\beta_{i}\right)^2 ,
\end{split}
\label{eq:baseline-response}
\end{equation}
where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the environmental gradient, respectively. Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods). For the sake of simplicity, we use only one environmental variable to characterize the                                                                             species' probability distribution. That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
\begin{figure}[ht]
  \centering
    \includegraphics[width=1\textwidth]{figures/all-distributions}
	   \caption{Different response curves. Panel (a) shows the probability density function characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\gamma$, when $\mu=0$. Panel (b) shows the probability density function characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\nu$, when $\gamma=1$ and $\mu=0$. Panel (c) shows the probability density function characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\lambda$, when $\gamma=1$ and $\mu=0$. Panel (d) shows the probability density function characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\lambda$ and $\nu$, when $\gamma=1$ and $\mu=0$. Notice that for each case, we chose $\alpha$ values that normalize the probability distributions.}
      \label{fig:response}
\end{figure}

\subsubsection*{Model priors}
The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the floristic database. To do so, we define the prior distributions for the parameters in model (\ref{eq:baseline-response}) as:
\begin{equation} 
\begin{split}
\beta_{i}  & \sim \text{MVNormal}\left(\hat{\beta}, \Sigma^{\beta}\right)\\
\text{log}(\gamma_{i})  & \sim \text{MVNormal}\left(\hat{\gamma}, \Sigma^{\gamma}\right)\\
\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
\hat{\beta}, 
\hat{\gamma}, 
\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
\sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
\end{split}
\label{eq:baseline-priors}
\end{equation} 
where parameters $\gamma_i$ and $\beta_i$ are expressed as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta}$ and $\Sigma^{\gamma}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively. We define these variance-covariance matrices as follows:
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance-baseline}
\end{equation}
where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$, and $\delta_{ij}$ is the Kronecker delta. Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information. In the floristic database, this information is represented by the set of ordinal traits specified for the different species. While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}). In each covariance matrix, the hyperparameter $\rho$ determines the rate of decline of the covariance between any two species, and $\eta$ defines its maximum value. The hyperparameter $\sigma$ describes the additional covariance between the different observations for any given species. For all these hyperparameters, we choose weakly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.

\subsubsection*{Alternative variance-covariance structures}
The model structure defined above allows us to test how different sources of information characterize each of the different parameters. Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}). For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits. One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\sum_k\rho_{k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance-complex}
\end{equation}
where now $\rho_{k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\beta_i$.

@@

\subsection*{Modifying the baseline model}
We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system. Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and prior information. We focused on two properties: fat-tailed and skewed responses.

@@

One way to test the extent to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes. That is, we can consider the prior distribution for these parameters as multivariate normal with a variance covariance matrix $\Sigma$. As described before, this variance-covariance matrix can be estimated following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are. Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.

>>>>

%Title paper
\title{\vspace{-1cm}
Model linearity breeds contempt: using Bayesian non-linear models to uncover broad macroecological patterns}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
% Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[2,3]{\normalsize Antoine Guisan}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
\affil[3]{\footnotesize Institute of Earth Surface Dynamics, University of Lausanne, Lausanne, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

\section*{Abstract}
Species' realized niches are classically pictured as bell-shaped probability distributions. These distributions, however, can actually take many different forms. For example, fat-tailed or skewed responses are very common across fields, as these can naturally emerge as a result of several ecological processes. While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies have shaped ecological communities. Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover such properties. These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies. With this approach, we are able to distil the shape of empirical plant distributions, which helps us tackle long-standing hypotheses regarding the way ecological communities are assembled across space. In particular, we study the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients. Moreover, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical contexts. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

@@

%############# NEW INTRO ##############
%Bell shape is a good option, but it is unclear what the real shape is
One of the central goals of ecology is to understand the ways species are distributed across space and time. While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, sagarinMovingAssumptionsUnderstand2006}. In practice, there is a strong argument to be made in favour of assuming these to be bell shaped. Namely, if all that we are willing to assume about species' distributions is that these occupy finite geographic ranges, the most conservative statistical approach is to model their distribution as Gaussian (i.e.~the corresponding maximum entropy distribution; \citealt{frankCommonPatternsNature2009}). That said, there is currently no general agreement on the basic shape of species' realized niches. Indeed, many factors can play a role in defining their shape, and several natural processes can lead to non-normal distributions.

@@

%they can also reveal differences among species regarding the way they assemble along a gradient and broad macroecological patterns defining natural constraints shaping communities. 
%The general properties of species' distributions are valuable because they point to important ecological processes shaping species' realized niches. 
%For example, a relationship or lack thereof between range size and elevation would speak to classical ecological hypotheses (i.e.~Rapopor's rule;  \citealt{stevensElevationalGradientAltitudinal1992}) and could hint at the existence of general biogeographical constraints that shape species' distributions along gradients.
%Moreover, properties such as the skewness of species' distributions might capture other underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, channellDynamicBiogeographyConservation2000}. For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients. This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments \citep{linderGeographicRangeStructure2000}. That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, helmuthBIOPHYSICSPHYSIOLOGICALECOLOGY2004, siefertHowClimateDispersal2015}. Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem to solve. Indeed, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species along environmental gradients.
%Understanding the shape of species' realized niches and the extent to which these vary across species is therefore a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
%Understanding the shape of species' realized niches and the extent to which these vary across species is however challenging statistical problem. That is, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.
%With all of that being said, measuring these properties is a challenging statistical problem, and we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.

Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000}. These models are fundamental to the scientific community for predicting changes in species' geographic distributions and the effects of environmental disturbances. Such frameworks, however, commonly assume an underlying linear relationship between covariates (but see `semiparametric models'; \citealt{norbergComprehensiveEvaluationPredictive2019}). This is useful because it simplifies the optimization process, but it might not be ideal when studying and comparing the shape of species' distributions along environmental gradients. First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified. While this might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}. Second, the aforementioned structural constraints also limit our ability to include any prior information to our parameter estimates. Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}). Finally, some have proposed several non-linear structures to characterize several features of individual species' response curves \citep{huismanHierarchicalSetModels1993}. Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following most of these model structures, these are generally not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g. sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).

In this work, we rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy. In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns. Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or neophytes, might respond to the environment. More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

@@

Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 10 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). In addition to species' environmental preferences, the floristic data also contains information on species introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades). We describe this information in more detail in the Supplementary Table 1.

%On the other hand, the information regarding species' ecological strategies represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology and life-history of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). Finally, and i

@@

\subsubsection*{Response curve}
To choose an appropriate response curve, we first need to agree on what we truly know about the system. Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance. While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions. Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution (Fig~\ref{fig:response}a). That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we can define species' responses to the environment as
\begin{equation}
\begin{split}
y_{ij} & \sim F\left(p_{ij}\right) \\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \gamma_{i} \left(x_{j}-\beta_{i}\right)^2 ,
\end{split}
\label{eq:baseline-response}
\end{equation}
where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the environmental gradient, respectively. Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods). For the sake of simplicity, we use only one environmental variable to characterize the                                                                             species' probability distribution. That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
\begin{figure}[h]
  \centering
    \includegraphics[width=1\textwidth]{figures/all-distributions}
	   \caption{Different response curves. Panel (a) shows the distribution shapes characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\beta$, $\alpha$ and $\gamma$. Panel (b) shows the distribution shapes characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\beta$, $\alpha$ and $\nu$, when $\gamma=1$. Panel (c) shows the distribution shapes characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\beta$, $\alpha$ and $\lambda$, when $\gamma=1$. Panel (d) shows the distribution shapes characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\beta$, $\alpha$, $\lambda$ and $\nu$, when $\gamma=1$. Notice that for all panels, we chose $\alpha$ values such that $\text{p}_i = \exp(-\alpha_i)$.
	   }
      \label{fig:response}
\end{figure}

\subsubsection*{Model priors}
The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the floristic database. To do so, we define the prior distributions for the parameters in model (\ref{eq:baseline-response}) as:
\begin{equation} 
\begin{split}
\beta_{i}  & \sim \text{MVNormal}\left(\hat{\beta}, \Sigma^{\beta}\right)\\
\text{log}(\gamma_{i})  & \sim \text{MVNormal}\left(\hat{\gamma}, \Sigma^{\gamma}\right)\\
\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
\hat{\beta}, 
\hat{\gamma}, 
\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
\sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
\end{split}
\label{eq:baseline-priors}
\end{equation} 
where parameters $\gamma_i$ and $\beta_i$ are expressed as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta}$ and $\Sigma^{\gamma}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively. We define these variance-covariance matrices as follows:
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance-baseline}
\end{equation}
where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$, and $\delta_{ij}$ is the Kronecker delta. Such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information. In the floristic database, this information is represented by the set of ordinal traits specified for the different species. While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}). In each covariance matrix, the hyperparameter $\rho$ determines the rate of decline of the covariance between any two species, and $\eta$ defines its maximum value. The hyperparameter $\sigma$ describes the additional covariance between the different observations for any given species. For all these hyperparameters, we choose weakly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$. Notice that other structures can be used to define the covariance matrices of the different Gaussian processes \citep{mcelreathStatisticalRethinkingBayesian2020}, including structures that account for multiple distance matrix $D_{ij}$ for any given parameter.

@@

\subsection*{Modifying the baseline model}
%We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system. Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and prior information. We focused on two properties: fat-tailed and skewed responses.

@@

%One way to test the extent to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes. That is, we can consider the prior distribution for these parameters as multivariate normal with a variance covariance matrix $\Sigma$. As described before, this variance-covariance matrix can be estimated following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are. Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.

\subsection*{Evaluating the log-likelihood}
One way to understand the shape of distributions is by evaluating the computed log-likelihood values across the environmental gradient. Comparing these values, one can understand what aspects of the shape of distributions are missing. To do so, for every sample of the model, we computed the log-likelihood values and the normalized probability distribution. This normalized probability is defined such that its maximum is set to 1 for all species in our dataset. In particular, for a heavy-tailed and skewed response, the normalized probability distribution was calculated for every sample of the Bayesian model using Eq.~(\ref{eq:heavy-skewed-response}), where $\alpha_i$ was set to 0 for any value of $x_{j}$. Notice that the normalized probability distribution is interesting when comparing the log-likelihood values across species because it can be used to understand whether the model errors are at the tails of the distributions or their center.

\section*{Results}
We studied the distribution data to characterize species' realized niches along the main axis of variation of all environmental variables. Using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation. This allowed us to map the center and variance of species' distributions along the environmental gradient (Fig.~\ref{fig:correlation}).  Studying the relationship between these properties, we found these to be negatively correlated (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model were positively correlated; Fig.~\ref{fig:correlation}). This means that species found at lower elevations have generally wider distributions than those at higher elevations. The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~XX) as well as when using ordinal data (Supplementary Fig.~XX); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~XX). The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships. In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and variance, respectively (i.e. $\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~XX). This implies that, at higher elevations, species' distributions generally have lower amplitudes.
